Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale

### 2024-03-06 paper/main.tex
small modifs

It shall be noted that this dependency of the slumping velocity on the bottom slope exists despite the slumping equilibrium being unmodified by the presence of a slope-induced acceleration (see section~\ref{sec:intro} for details). As suggested in \citet{Gadal2023}, the latter acceleration may instead affect the complicated dynamics of the early transient acceleration phase, involving strong vertical motions and significant interfacial friction \citep{Cantero2007}.
%
One can write an energetic balance for the entire current between its initial state and the end of this early acceleration phase \citep{Gadal2023}
\begin{equation}
	\label{eq:energetic_balance}
	\left[\frac{1}{2}\rho_{0}u_{\text c}^{2} - B \Delta\rho g \sin\alpha L\right] - A \Delta\rho g \cos\alpha h_{0} =  - \frac{1}{2}c_{\text d}\rho_{0}u_{\text c}^{2} \frac{L}{h_{0}},
\end{equation}
where $A$ and $B$ are scaling constants, $c_{\text d}$ is a drag coefficient typically on the order of $10^{-2}$ -- $10^{-3}$ \citep{hogg2004effects}, and $L$ accounts for the alongslope distance over which the current moves during this phase. \citet{Cantero2007} found it to be independent of the lock aspect ratio or initial buoyancy, $L = 0.3 h_{0}$. Hence, we use here $L = D h_{0}$ with $D$ constant, which results in
\begin{equation}
	\label{eq:Fr_th}
	\mathcal{F}r = \frac{U_{\text c}}{U_0} = \frac{Fr_0}{\sqrt{1 + C_{\text D}}} \sqrt{1 + \frac{\tan\alpha}{S}},
\end{equation}
where $Fr_0$, $C_{\text D}$ and $S$ are new constants depending on $A$, $B$, $c_{\text d}$ and $D$. Note that $C_{\rm d}$ has little influence if $c_{\text d}$ is kept within the reasonable values mentioned previously, and $Fr_0$ is strongly restricted by the values in $\theta = 0^\circ$, such that the only truly adjustable parameter is $S$. As shown in Fig.~\ref{fig:fig4}, this simple dimensional model is able to reproduce the observed variation of the dimensionless velocity in the bottom slope in the range [$5^\circ$, $35^\circ$] while keeping the scaling constants within reasonable values.
%
For $\alpha < 5^\circ$, experimental data points are significantly scattered, and \eqref{eq:Fr_th} almost systematically overpredicts the $\mathcal{F}r$ values. A plausible explanation for this discrepancy could be the difference in the width-to-depth ratio of the flumes (and associated wall dissipation) and in aspect ratios of the lock, which are not addressed in the energetic balance leading to \eqref{eq:Fr_th}.
%
For $\alpha > 35^\circ$, the measured dimensionless velocities ($\mathcal{F}r$) slowly stand above equation (\ref{eq:Fr_th}) curve. In this region of large slope angles, the equilibrium between the forces likely becomes a lot more in favor of the so-called slope effect than the pressure gradient. Consequently, the scaling constants may have to be adapted to this major change in the influence of the forces.

>>>>

It shall be noted that this dependency of the slumping velocity on the bottom slope exists despite the slumping equilibrium being unmodified by the presence of a slope-induced acceleration (see section~\ref{sec:intro} for details). As suggested in \citet{Gadal2023}, the latter acceleration may instead affect the complicated dynamics of the early transient acceleration phase, involving strong vertical motions and significant interfacial friction \citep{Cantero2007}.
%
One can write an energetic balance for the entire current between its initial state and the end of this early acceleration phase \citep{Gadal2023}
\begin{equation}
	\label{eq:energetic_balance}
	\left[\frac{1}{2}\rho_{0}u_{\text c}^{2} - B \Delta\rho g \sin\alpha L\right] - A \Delta\rho g \cos\alpha h_{0} =  - \frac{1}{2}c_{\text d}\rho_{0}u_{\text c}^{2} \frac{L}{h_{0}},
\end{equation}
where $A$ and $B$ are scaling constants, $c_{\text d}$ is a drag coefficient typically on the order of $10^{-2}$ -- $10^{-3}$ \citep{hogg2004effects}, and $L$ accounts for the alongslope distance over which the current moves during this phase. \citet{Cantero2007} found it to be independent of the lock aspect ratio or initial buoyancy, $L = 0.3 h_{0}$. Hence, we use here $L = D h_{0}$ with $D$ constant, which results in
\begin{equation}
	\label{eq:Fr_th}
	\mathcal{F}r = \frac{U_{\text c}}{U_0} = \frac{Fr_0}{\sqrt{1 + C_{\text D}}} \sqrt{1 + \frac{\tan\alpha}{S}},
\end{equation}
where $Fr_0$, $C_{\text D}$ and $S$ are new constants depending on $A$, $B$, $c_{\text d}$ and $D$. Note that $C_{\rm d}$ has little influence if $c_{\text d}$ is kept within the reasonable values mentioned previously, and $Fr_0$ is fixed by the values in $\theta = 0^\circ$ (also corresponding to previous literature), such that the only truly adjustable parameter is $S$. As shown in Fig.~\ref{fig:fig4}, this simple dimensional model is able to reproduce the observed variation of the dimensionless velocity in the bottom slope in the range [$5^\circ$, $35^\circ$] while keeping the scaling constants within reasonable values.
%
For $\alpha < 5^\circ$, experimental data points are significantly scattered, and \eqref{eq:Fr_th} almost systematically overpredicts the $\mathcal{F}r$ values. A plausible explanation for this discrepancy could be the difference in the width-to-depth ratio of the flumes (and associated wall dissipation) and in aspect ratios of the lock, which are not addressed in the energetic balance leading to \eqref{eq:Fr_th}.
%
For $\alpha > 35^\circ$, the measured dimensionless velocities ($\mathcal{F}r$) slowly stand above equation (\ref{eq:Fr_th}) curve. In this region of large slope angles, the equilibrium between the forces likely becomes a lot more in favor of the so-called slope effect than the pressure gradient. Consequently, the scaling constants may have to be adapted to this major change in the influence of the forces.
